---
author: Robert
date: '`r format(Sys.Date())`'
title: Cheyenne County Parcels
output:
  html_document:
    toc_float: false
    number_sections: true
    toc: true
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE) 

```



```{r}
library(data.table) # Use fread to import
library(tidyverse)
library(janitor)
library(tictoc)
library(stringr)
library(lubridate)
library(arrow)
library(here)
``` 


### Before you deply your regex expressions in your code, test them out with *regex testers*    
* [To install regexplain package by Garrick Aden-Buie](https://www.garrickadenbuie.com/project/regexplain/)  
* [rubular by Michael Lovitt](https://rubular.com)  
* [regex101 by Firas Dib](https://regex101.com/)    
* [regexr by Grant Skinner](https://blog.gskinner.com/archives/2008/03/regexr_free_onl.html)    
* [regexr - regexr.com](https://regexr.com/)   
* [regexpal - part of "Dan's Developer Tools]  






### Import Cheyenne County Parcel data (Ownership)  
* Sales_Date as date does not import
* Sales_Date as character *does* import but not all of the date obs convert to date type.  
* For consistent date conversion, convert the excel file to csv before import.  
* load with readlx = 0.45 seconds
* load with arrow = 0.01 seconds  

**Import parquet data with arrow** 
```{r}

tic()
parcels <- read_parquet(here("data/cheyenne_parcels.parquet"))
toc()
```


```{r eval=FALSE}  

tic()
# library(readxl)

# parcelpath <- "F:/D_Documents/GitHub-DasRotRad/regex_rfortherest/data/Cheyenne_Parcels.xlsx"
# 
# Cheyenne_Parcels <- read_excel(parcelpath, 
# # Cheyenne_Parcels <- read_excel("data/Cheyenne_Parcels.xlsx", 
#     col_types = c("text", "text", "text", 
#         "text", "text", "text", "numeric", 
#         "numeric", "numeric", "numeric", 
#         "numeric", "numeric", "text", "date", 
#         "numeric", "text", "text", "numeric", 
#         "numeric", "text", "numeric", "numeric", 
#         "numeric"))
# 
# 
# toc()
# View(Cheyenne_Parcels)
# 


# write_parquet(Cheyenne_Parcels, here("data/cheyenne_parcels.parquet"))

# write_csv_arrow(cars, "cars.csv")

# write.csv(Cheyenne_Parcels, "./data/Cheyenne_Parcels.csv")
```  


### Import Parcel data, Sales Date as Character 
* Chunk not evaluated  
* import takes 0.27 seconds  
* has sales_date  
* fast but I find problems coverting number to as.Date()   

```{r eval=FALSE}
library(readxl)
tic()
Cheyenne_Parcels_c <- read_excel("data/Cheyenne_Parcels.xlsx") 
toc()
View(Cheyenne_Parcels)
```


```{r eval=FALSE}
# library(xlsx)
library(writexl)

write_xlsx(Cheyenne_Parcels,path=here("data/cheyenne_owners.xlsx"))

```


### Import Parcels from csv with *fread()*  from data.table package 


* import takes 0.04 seconds  
* class(parcels0$SALES_DATE) SALES_DATE was imported as a character with fread.
* convert character string to date object.
* increases import from 0.04 to 0.06 seconds. 
* clean_names from Janitor package  


### Importing with fread does not read dates.   

* Chunk not evaluated  

```{r eval=FALSE}
tic()
parcels0 <- fread("./data/Cheyenne_Parcels.csv") %>% 
  mutate(SALES_DATE = as_date(SALES_DATE)) %>% 
  clean_names()
toc()
```
* class(parcels0$SALES_DATE) 

### problems 
* When using *fread* to import from xlsx files, dates were corrupted.
* could not convert all to date type
```{r eval=FALSE}
parcels <- Cheyenne_Parcels %>%  
  clean_names() %>% 
  select(parcel_id, current_ow, street_add, legal_desc, total_asse, sales_date, property_t)  
  # as.POSIXct(sales_date, "ymd")
 #  as.Date(sales_date, format = "%Y.%B.%d")

  
parcels
```

## Import from CSV using library(readr)  
* Chunk not evaluated  
```{r eval=FALSE}
library(readr)


tic()
Cheyenne_Parcels <- read_csv("data/Cheyenne_Parcels.csv", 
    col_types = cols(SALES_DATE = col_date(format = "%m/%d/%Y")))
toc()
View(Cheyenne_Parcels)
```

### Clean Names  *trim_ws = TRUE*  
* trim white space.  Print in console.  If any white spaces there will be quotes. 

```{r}

parcels <- parcels %>% 
  clean_names() %>% 
  select(parcel_id, current_ow, street_add, street_add, legal_desc, total_asse, sales_date, shape_area)
  trim_ws = TRUE  
  
```  


### Look for White space
```{r}
parcels %>% 
  mutate(leadingst = str_detect(current_ow, "^ "),
         trailingws = str_detect(current_ow, " $")) %>% 
  select(leadingst, trailingws)  
#  filter(leadingst=="TRUE")
#  filter(trailingws=="TRUE")

# remove whitespaces
# parcels %>% 
  # mutate(current_ow = str_squish(current_ow)) %>% 
  # select(current_ow)

```

```{r}
parcels  %>% 
  select(street_add) %>% 
  head(15)
# tidyr::separate(current_ow, "owner2", "address", ",")

``` 


### Splitting Columns with tidyr::separate()  
* [Joachim Schork's Statics Globe](https://www.youtube.com/watch?v=xrrWLKRfZWY)  

```{r}
str_split(parcels$street_add, pattern = "\\s[A-Z]{2}$", 2) %>% 
  head(15)
```


```{r} 

parcels0 <- parcels  %>% 
  clean_names() %>% 
  select(street_add) %>%
  separate(street_add, c("col1", "address", "city_state", "zipcode"), ",") %>% 
  mutate(across(where(is.character), str_trim))
  view()
  
``` 

### compare dfs
```{r}

all_equal(parcels, parcels0)
compare_df_cols(parcels, parcels0)


```


### Splitting Column by number of characters  
```{r}
parcels0 %>% 
 # data.frame() %>% 
  str_split(string = parcels0$city_state, pattern = "\\s..$") %>% 
  sample_n(6)
```


```{r}
separate(parcels0, city_state, into=c("city", "state"), sep = regex("\\s[A-Z]{2}$"))


  separate(parcels0$city_state, into = c("city", "state"), sep = " ")
  
x1 <- str_split(parcels0$city_state, pattern = ("\\s[A-Z]{2}$"), n=2, simplify = TRUE)
x1 <- str_split(parcels0$city_state, pattern = " ", simplify = TRUE)

```

```{r}
parcels %>% 
  sample_n(20) %>% 
  select(city_state) %>% 
separate(parcels, city_state, into=c("city", "state"), sep = "\\w")

library(treemapify)
ggplot(df_roy, aes(area = TotalRoyAc,
        fill = initials,
        label = TotalRoyAc,
        subgroup = initials)) +
  geom_treemap() +
  geom_treemap_subgroup_border() +
  geom_treemap_subgroup_text(place = "centre",
              grow = F,
              alpha = 0.5,
              colour = "black",
              fontface = "italic",
              min.size = 0) +
  geom_treemap_text(colour = "white", place = "topleft", reflow = T) +
      labs(y = " ",
      x = NULL,
      title = "Who is Contributing the Greatest Number of Royalty Acres?",
      subtitle = "Royalty Acres = Sum of Roy Acres * NRI")
```


```{r}
x <- parcels %>% 
  sample_n(10) %>% 
  mutate(city_state = str_squish(city_state)) %>% 
  trimws(city_state)
# x<-  trimws(x$city_state) 
  
#  x  <- pillar::trim_ws(x$city_state,which="both")

strsplit(x, split= " ")
tibble(x)
```

### Create a dataframe for dput examples  

```{r}
library(splitstackshape)

class(city_state) 




cSplit("cityDT", sep = " ", drop = FALSE) # separate animals_name column, but keep it

```

